Bird Audio Detection Challenge

### 2017-03-20 paper/main.tex
Fixed many grammatical mistakes

	Acoustic detection of birds in the wild is useful in many biological
	scenarios, including estimation of bird population and species-specific
	acoustic monitoring. Most of current methods are either manual or
	semi-automatic, requiring labor-intensive processing/tuning. We hereby
	present three alternative methods, as part of the Bird Audio Detection
	Challenge, hosted by Queen Mary University of London. Two methods are our
	own submission, with the third one as the winning solution submitted by
	Thomas Grill. All methods utilize deep neural networks for automatic,
	tuning-free bird presence detection, with the first two methods
	emphasizing network architectures, and the third one utilizing multiple
	signal preprocessing techniques.

@@

In response to the need for automatic, tuning-free bird presentce detection
algorithm, \textit{Bird Audio Detection Challenge} was created, hosted by
Queen Mary University of London and other universities. Three datasets were
released as part of the challenge, and teams across the world compete for
the best performance on the task of bird audio detection.

In section 2, we discuss some of the competition details, including the
datasets used, task description and the evaluation metrics. Section 3
presents three methods for bird acoustic detection, two of which are our
submission to the challenge, and the third one being the best performing
method submitted by Thomas Grill. Section 4 discusses some experiments and
techniques used to improve the peformance. Section 5 displays the
evaluation results of the three methods, and section 6 presents future
possible work from this challenge.

@@

For the challenge, teams have to design a system that takes a short audio
clip, and returns a binary label denoting the presence/absence of bird
sound of any species. Teams are also encouraged to output a floating-point,
probability score in the range of 0 to 1 for evaluation purposes.

@@

\textbf{Warblr:} Crowd-sourced project with many smartphone recordings from
around the United Kingdom. Very diverse in sound categories, even including
human bird imitations.

@@

Our first method resembles basic characteristics of SoundNet by Aytar et
al.\cite{soundnet} and Residual Net by He et al.\cite{residual_net}. The
network is fully convolutional, with one-dimensional convolutions followed
by non-linearities (ReLU) layers.  This network is suitable for audio
signals, as it is invariant to translation and allows stacking of
convolutional layers for detecting high-level features.

Network also has shortcuts from earlier layers of the network to later
stages, allowing easier gradient back-propagation to early layers of the
network. See Figure 1 for network details.

@@

Our second method utilizes a series of pooling and convolutional layers to
extract features from the input audio at different temporal resolutions,
and concatenate them together before some final convolutional layers.

Network can be divided into three stages. In the first stage, we first
create two copies of the input, one reshaped to 100,000 by 4, and the other
reshaped to 200,000 by 2. \(x_{100,000*4}\) is then passed through a single
convolutional layer while \(x_{200,000*2}\) is passed through a series of 5
averaging pooling layers, followed by a convolutional layer at the end of
each pooling layer. Note here that the input of one pooling layer is the
output of the previous pooling layer, the outputs of the convolutional
layers serve as features extracted at input of different temporal
resolution.

In the second stage, we combined the outputs of the 6 convolutional layers
in Stage 1 into one feature vector. First, we find the maximum for each
convolutional output. We denote them as the energy for each convolutional
layer. We then concatenate them with their corresponding convolutional
output, pass each concatenated feature vector through a pooling layer of
specified size, and concatenate all 6 pooled feature vectors into one. Due
to different lengths of output from the 6 convolutional layers in stage 1,
we pass them through pooling layers of different sizes to ensure that they
are of the same dimension for concatenation.

In the final stage, we pass the output from Stage 2 through a batch
normalization layer, followed by 7 additional convolutional layers, with a
\(max()\) operation between the 4th and 5th layer. Final output is then
flattened into size (1,2), ready to be used in a softmax classification
loss function.

@@

% TODO: Explain that spectrograms were used during training

This method is the winning algorithm to the challenge, submitted by Thomas
Grill. The neural network uses processed spectrograms of audio clips as
input, instead of a one-dimensional vector of signals.The network consists
of 4 convolutional layers with pooling, and 2 dense layers. Multiple signal
pre-processing techniques are used, including silence/noise trimming, and a
different data augmentation technique involving shifting the spectrograms
in time and frequency.  Pseudo-labeling and bagging techniques are also
deployed, which we will discuss in section 4.

@@

During our development, we experimented with various pre- and
postprocessing techniques, as well as some training strategies.

@@

The given labeled data from both datasets is split into 10 folds. Within
each fold, number of positive(bird present) and negative(bird absent)
samples are balanced. During training, number of samples loaded from each
datasets are also balanced, to ensure no bias.

@@

For method 3, a brief description and code base is available online
at\cite{winning_blog}

@@

We deploy data augmentation as our only pre-processing technique for method
1 and 2. Given one base sound clip, positive or negative (bird
present/absent), and another negative sound clip, we concatenate them along
the y-axis, with the base clip weighted by a factor \(r\) between 0.25 and
1, and the negative sound clip weighted by a factor \(1-r\). So we have:
\begin{equation}
	x_{augmented}=r*x_{base} + (1-r)*x_{negative}
\end{equation}
If \(r\) equals to 1, then there would be no augmentation, and if \(r\)
equals to 0.25, only 25\% of the base sound clip will be preserved. We
sample \(r\) on a positively skewed interval (0.25,1) to ensure bird sounds
in positive labels are not diminished.

Label for the augmented clip will remain the same as the base clip, as the
augmentation is an \textit{or} process, preserving the original label when 
\textit{or} with a negative label.

@@

One main difficulty we faced is the small size of training data. With no
regularization, our networks consistently over-fit to the training data,
perform near perfect on training and our own held-out test data, but
perform poorly on the challenge data. To compensate this, we reduce the
capacity of our models by reducing the number of filters for each
convolutional layer. We do so until the network cannot perform perfectly on
the training and testing data.

@@

This post-processing technique averages prediction scores from many
sessions. Due to the random cropping of audio signals during the data
loading phase, different evaluation of the same audio clip will yield
slightly different results. Therefore we predict 100 sets of scores for the
held-out challenge dataset, and average them to capture the true scores for
submission.

@@

This technique is used by many contestants and attempted by us. The
assumption is that high-confidence predictions on the held-out challenge
dataset are accurate enough to be used as extra data for retraining or
fine-tuning. Basic procedures are:

@@

\(r\) can be obtained through manual spot-checking of the predictions. Too
much time spend on finding the threshold will defeat the purpose of the
challenge.

@@

For method 1 and 2, no signal pre-processing techniques were deployed,
resulting in very noisy inputs to our network. It will be interesting to
find out if pre-processing will aid the prediction accuracy of our two
networks.

@@

Bird acoustic detection is an important task for many biological
applications, but the massive amount of data available makes it difficult
to extract any useful information from it. The Bird Audio Detection
challenge has advanced this field in many ways, and many excellent
automatic, tuning-free bird acoustic detection algorithms are introduced.
We presented three methods, first two with a focus on unique network
architecture, and the third one as the embodiment of many years of machine
listening research.  Many topics are still open for research, including
many tasks that already have a solution, but only achieved through manual
or labor intensive.

>>>>

	Acoustic detection of birds in the wild is useful in many biological
	scenarios, including estimation of bird population and species-specific
	acoustic monitoring. Most current methods are either manual or
	semi-automatic, requiring labor-intensive processing/tuning. We hereby
	present three alternative methods, as part of the Bird Audio Detection
	Challenge, hosted by Queen Mary University of London. The first two
	methods are of our own submission, the third being the winning solution
	submitted by Thomas Grill. Each of said methods utilize deep neural
	networks for automatic, tuning-free bird presence detection: the first
	two emphasize network architectures, while the third utilizes multiple
	signal preprocessing techniques.

@@

In response to the need for an automatic, tuning-free bird presence
detection algorithm, \textit{Bird Audio Detection Challenge} was created,
hosted by Queen Mary University of London and other universities. Three
datasets were released as part of the challenge, and teams across the world
compete for the best performance on the task of bird audio detection.

In section 2, we discuss some of the competition details, including the
datasets used, task description and the evaluation metrics. Section 3
presents three methods for bird acoustic detection, two of which are our
submission to the challenge, and the third one being the best performing
method submitted by Thomas Grill. Section 4 discusses some experiments and
techniques used to improve the peformance. Section 5 displays the
evaluation results of the three methods, and section 6 presents future
possible work that may result from this challenge.

@@

For the challenge, teams have to design a system that takes a short audio
clip, and returns a binary label denoting the presence/absence of any bird
species' sound. Teams are also encouraged to output a floating-point,
probability score in the range of 0 to 1 for evaluation purposes.

@@

\textbf{Warblr:} Crowd-sourced project with many smartphone recordings.
Very diverse in sound categories, even including human bird imitations.

@@

% TODO: Weird spacing issue in first and second sentence

Our first method resembles basic characteristics of SoundNet by Aytar et
al.\cite{soundnet} and Residual Net by He et al.\cite{residual_net}. The
network is fully convolutional, with one-dimensional convolutions followed
by non-linear (ReLU) layers. This network is suitable for audio
signals, as it is invariant to translation and allows stacking of
convolutional layers for detecting high-level features.

This network also has shortcuts from earlier layers of the network to later
stages, allowing easier gradient back-propagation to early layers of the
network. See Figure 1 for network details.

@@

Our second method utilizes a series of pooling and convolutional layers to
extract features from the input audio at different temporal resolutions,
and concatenate them before some final convolutional layers.

This network can be divided into three stages. In the first stage, we first
create two copies of the input, one reshaped to 100,000 by 4, and the other
reshaped to 200,000 by 2. \(x_{100,000*4}\) is then passed through a single
convolutional layer while \(x_{200,000*2}\) is passed through a series of 5
averaging pooling layers, followed by a convolutional layer at the end of
each pooling layer. Note here that the input of one pooling layer is the
output of the previous pooling layer; the outputs of the convolutional
layers serve as features extracted at input of different temporal
resolution.

In the second stage, we combine the outputs of the 6 convolutional layers
in stage 1 into one feature vector. First, we find the maximum for each
convolutional output. We denote them as the energy for each convolutional
layer. We then concatenate them with their corresponding convolutional
output, pass each concatenated feature vector through a pooling layer of
specified size, and concatenate all 6 pooled feature vectors. Due to
different lengths of output from the 6 convolutional layers in stage 1, we
pass them through pooling layers of different sizes to ensure that they are
of the same dimension for concatenation.

In the final stage, we pass the output from stage 2 through a batch
normalization layer, followed by 7 additional convolutional layers, with a
\(max()\) operation between the 4th and 5th layer. Final output is then
flattened into size (1,2), ready to be used in a softmax classification
loss function.

@@

This method is the winning algorithm to the challenge, submitted by Thomas
Grill. The neural network uses processed spectrograms of audio clips as
input, instead of a one-dimensional vector of signals. The network consists
of 4 convolutional layers with pooling, and 2 dense layers. Multiple signal
pre-processing techniques are used, including silence/noise trimming, and a
different data augmentation technique involving shifting the spectrograms
in time and frequency.  Pseudo-labeling and bagging techniques are also
deployed, which we will discuss in section 4.

@@

During our development, we experimented with various pre- and
post-processing techniques, as well as some training strategies.

@@

The given labeled data from both datasets is split into 10 folds. Within
each fold, the number of positive(bird present) and negative(bird absent)
samples are balanced. During training, the number of samples loaded from
each datasets are also balanced, to ensure no bias.

@@

For method 3, a brief description and code base is available online
at\cite{winning_blog}.

@@

We deploy data augmentation as our only pre-processing technique for method
1 and 2. Given one base sound clip, positive or negative (bird
present/absent), and another negative sound clip, we concatenate them along
the y-axis, with the base clip weighted by a factor \(r\) between 0.25 and
1, and the negative sound clip weighted by a factor \(1-r\). Thus we have:
\begin{equation}
	x_{augmented}=r*x_{base} + (1-r)*x_{negative}
\end{equation}
If \(r\) equals to 1, then there would be no augmentation, and if \(r\)
equals to 0.25, only 25\% of the base sound clip will be preserved. We
sample \(r\) on a positively skewed interval (0.25,1) to ensure bird sounds
in positive labels are not diminished.

Label for the augmented clip will remain the same as the base clip, as the
augmentation is an \textit{or} process, preserving the original label when 
"\textit{or}-ed" with a negative label.

@@

One main difficulty we faced was the small size of training data. With no
regularization, our networks consistently over-fit to the training data,
performed nearly perfectly on training and our own held-out test data, but
performed poorly on the challenge data. To compensate for this, we reduced
the capacity of our models by reducing the number of filters for each
convolutional layer. We did so until the network could not perform
perfectly on the training and testing data.

@@

This post-processing technique averages prediction scores from many
sessions. Due to the random cropping of audio signals during the data
loading phase, different evaluation of the same audio clip will yield
slightly different results. Therefore we predicted 100 sets of scores for the
held-out challenge dataset, and averaged them to capture the true scores for
submission.

@@

This technique is used by many contestants and was attempted by us. The
assumption is that high-confidence predictions on the held-out challenge
dataset are accurate enough to be used as extra data for retraining or
fine-tuning. Basic procedures are:

@@

\(r\) can be obtained through manual spot-checking of the predictions. Too
much time spent on finding the threshold will defeat the purpose of the
challenge.

@@

For methods 1 and 2, no signal pre-processing techniques were deployed,
resulting in very noisy inputs to our network. It will be interesting to
find out if pre-processing will aid the prediction accuracy of our two
networks.

@@

Bird acoustic detection is an important task for many biological
applications, but the massive amount of data available makes it difficult
to extract any useful information from it. The Bird Audio Detection
challenge has advanced this field greatly--many excellent automatic,
tuning-free bird acoustic detection algorithms are introduced.  We
presented three methods, the first two with a focus on unique network
architectures, and the third one as the embodiment of many years of machine
listening research.  Many topics are still open for research, including
numerous tasks that already have a solution, but can only be achieved
through manual or labor intensive annotation.
